Parameter-less Population Pyramid

### 2013-12-17 Parameterless Population Pyramid/main.tex
First revision pass complete.

There have been other model based optimization techniques that use little to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
In general, LTGA is the most powerful general optimizer with few parameters.

@@

@@

In total, our experiments performed 100 runs of each optimization algorithm across
7 benchmark problems, with each run limited to 100 million evaluations.
In performing data collection, we performed over 150,000 runs, totaling approximately
1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
set the population size of LTGA.  Note that these runs should be considered a training phase,
as such all reported results for LTGA use independent runs to create a testing phase.
This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
advantage over the others in that it has pre-learned the correct population size.
All of the code used to perform our experiments
as well as individual run results and analysis scripts are available from our
website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.

Figure~\ref{fig-results} does provide one misleading
exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears
level out at high genome sizes.  This is entirely an artifact of the extremely high
variance of this algorithm on this problem.  Note that many of the large problem
sizes failed to find the optimum for the majority of runs (no dot drawn for that size).

@@

Table~\ref{table-results} provides a close up look at the largest problem sizes
in which LTGA was able to find the global optimum reliably in less than 100 million
evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
in p-values less than $10^-22$, with the exception of Deceptive Step Trap which
resulted in $0.00025$.  As such these results are highly significant.  The amount
of speedup ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
cases P3 outperformed LTGA and all other comparison techniques.

@@

Yet unlike many theory based algorithms, we believe P3 can achieve bounds on more
complex and interesting landscapes.  Consider a sketch for a proof on the Deceptive
Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
random individual and optimize it to a local optima using the hill climber.  In
$P_0$ all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters
will move only complete traps between solutions, and once there is sufficient
diversity in the population there will be a crossover for each individual trap.
When performing a crossover of a single trap there can only be a single alternative
value, meaning if the candidate solution has all 0s the donor must have all 1s, or vice
versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
individual and has a crossover that moves that trap, all solutions produced by crossing
$P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
all traps we can say that once any $P_i$ has at least one solution with the optimized
version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
The only remaining portion of the proof is to determine how many evaluations are required to reach this point,
which should be relatively trivial.

>>>>

There have been other model based optimization techniques that use little to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
%In general, LTGA is the most powerful general optimizer with few parameters.

@@

In total, our experiments performed 100 runs of each optimization algorithm across
7 benchmark problems, with each run limited to 100 million evaluations.
In performing data collection, we performed over 150,000 runs, totaling approximately
1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
set the population size of LTGA.  Note that these runs should be considered a training phase,
and as such all reported results for LTGA use independent runs to create a testing phase.
This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
advantage over the others in that it has pre-learned the correct population size.
All of the code used to perform our experiments
as well as individual run results and analysis scripts are available from our
website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.

Figure~\ref{fig-results} does provide one misleading
exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears to
level out at high genome sizes.  This is entirely an artifact of the extremely high
variance of this algorithm on this problem.  Note that many of the large problem
sizes failed to find the optimum for the majority of runs (no dot drawn for that size).

@@

@@

Table~\ref{table-results} provides a close up look at the largest problem sizes
in which LTGA was able to find the global optimum reliably in less than 100 million
evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
in p-values less than $10^{-22}$, with the exception of Deceptive Step Trap which
resulted in $0.00025$.  As such these results are highly significant.  The amount
of speedup ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
cases P3 outperformed LTGA and all other comparison techniques.

@@

Yet unlike many theory based algorithms, we believe P3 can achieve bounds on more
complex and interesting landscapes.  Consider a sketch for a proof on the Deceptive
Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
random individual and optimize it to a local optima using the hill climber.  In
$P_0$ each trap in all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters
will move only complete traps between solutions, and once there is sufficient
diversity in the population there will be a crossover for each individual trap.
When performing a crossover of a single trap there can only be a single alternative
value, meaning if the candidate solution has all 0s the donor must have all 1s, or vice
versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
individual and has a crossover that moves that trap, all solutions produced by crossing
$P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
all traps we can say that once any $P_i$ has at least one solution with the optimized
version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
The only remaining portion of the proof is to determine how many evaluations are required to reach this point,
which should be relatively trivial.
